---
title: "Getting started with dipca"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dipca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

## Why dipca?

Standard PCA finds directions of maximum variance, but it ignores temporal
structure. When your data are multivariate time series, the most *interesting*
directions are often those that are **predictable from their own past** --
not merely high-variance. The dipca package provides three methods that extract
such temporally structured latent components:

| Method | Function | Objective |
|--------|----------|-----------|
| DiPCA  | `dipca()` | Maximize autocovariance of latent scores |
| DiCCA  | `dicca()` | Maximize canonical correlation between scores and their AR prediction |
| DiPLS  | `dipls()` | Maximize dynamic cross-covariance between an input block X and output block Y |

All three return objects that integrate with the
[multivarious](https://github.com/bbuchsbaum/multivarious) framework, so you
get `scores()`, `project()`, `reconstruct()`, `predict()`, and `residuals()`
out of the box.

```{r load-packages}
library(dipca)
library(multivarious)
```

## Quick example

Simulate a multivariate time series with two hidden dynamic components, fit
DiPCA, and recover them in a few lines:

```{r quick-example}
set.seed(42)
N <- 300; m <- 6

# Two latent AR(1) processes with different dynamics
t1 <- arima.sim(list(ar = 0.8), n = N)
t2 <- arima.sim(list(ar = 0.4), n = N)

# Mix into 6 observed variables + noise
P <- qr.Q(qr(matrix(rnorm(m * 2), m, 2)))
X <- cbind(t1, t2) %*% t(P) + matrix(rnorm(N * m, sd = 0.3), N, m)

# Fit DiPCA: extract 2 components with lag order 1
fit <- dipca(X, s = 1, l = 2, n_init = 3)

# Recovered scores
head(scores(fit))
```

The fitted object is a `bi_projector` from multivarious. You can project new
data, reconstruct, or forecast with familiar generic functions.

## DiPCA: unsupervised dynamic components

`dipca()` extracts components that maximize autocovariance -- a balance between
capturing variance and being temporally predictable.

```{r dipca-fit}
fit_pca <- dipca(X, s = 1, l = 2, n_init = 3, max_iter = 500)
```

**Scores** are the latent time series:

```{r dipca-scores, fig.height=5}
T_hat <- scores(fit_pca)

par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(T_hat[, 1], type = "l", col = "steelblue",
     ylab = "Score", main = "DiPCA Component 1")
plot(T_hat[, 2], type = "l", col = "firebrick",
     ylab = "Score", main = "DiPCA Component 2")
```

**Prediction** uses the fitted VAR model on the latent scores:

```{r dipca-predict}
pred <- predict(fit_pca, X)

# R-squared for each component
r2 <- sapply(1:2, function(j) {
  ok <- !is.na(pred$scores_hat[, j])
  cor(pred$scores[ok, j], pred$scores_hat[ok, j])^2
})
cat("Component R-squared:", round(r2, 3), "\n")
```

**Reconstruction** maps scores back to the original variable space:

```{r dipca-recon}
X_hat <- reconstruct(fit_pca)
cat("Reconstruction RMSE:", round(sqrt(mean((X - X_hat)^2)), 4), "\n")
```

**Residuals** provide a dynamic whitening filter -- the temporal structure is
removed:

```{r dipca-resid}
res <- stats::residuals(fit_pca, X)

# Compare autocorrelation before and after whitening
acf_before <- mean(sapply(1:m, function(j)
  acf(X[, j], lag.max = 1, plot = FALSE)$acf[2]))
acf_after <- mean(sapply(1:m, function(j) {
  ok <- is.finite(res$e_hat[, j])
  if (sum(ok) > 10) acf(res$e_hat[ok, j], lag.max = 1, plot = FALSE)$acf[2]
  else NA
}), na.rm = TRUE)

cat("Mean lag-1 ACF -- original:", round(acf_before, 3),
    " residuals:", round(acf_after, 3), "\n")
```

## DiCCA: components ordered by predictability

`dicca()` extracts components in descending order of predictability (R-squared).
This is useful when you want the *most dynamic* features first.

```{r dicca-fit}
fit_cca <- dicca(X, s = 1, l = 2, inner = "classic", n_init = 3)

cat("DiCCA R-squared per component:", round(fit_cca$R2, 3), "\n")
```

DiCCA also supports compact inner models (`inner = "ar"` or `"arma"`) that
fit AR/ARMA models per component instead of a joint VAR:

```{r dicca-ar}
fit_ar <- dicca(X, s = 1, l = 2, inner = "ar", n_init = 2)

# Compact models fit per-component AR/ARMA inner models; compute R² from predict()
pred_ar <- predict(fit_ar, X)
r2_ar <- sapply(1:2, function(j) {
  ok <- !is.na(pred_ar$scores_hat[, j])
  cor(pred_ar$scores[ok, j], pred_ar$scores_hat[ok, j])^2
})
cat("Compact DiCCA one-step R-squared:", round(r2_ar, 3), "\n")
```

An experimental `inner = "arima"` path exists for unit-root or complex dynamics;
enable it with `options(dipca.experimental_arima = TRUE)` before calling `dicca()`.

The same `predict()`, `residuals()`, and `reconstruct()` methods work for
both classic and compact variants.

## DiPLS: supervised dynamic components

When you have both input (X) and output (Y) blocks, `dipls()` finds dynamic
latent components that maximize the cross-covariance between X-scores and
Y-scores through a lagged inner model.

```{r dipls-fit}
# Simulate an output block driven by the first latent component
beta_true <- c(0.7, -0.3)
u <- stats::filter(t1, beta_true, sides = 1)
Y <- cbind(u, 0.5 * u + rnorm(N, sd = 0.2))
Y[is.na(Y)] <- 0

fit_pls <- dipls(X, Y, n_comp = 1, s = 1, mode = "fir", verbose = 0)

# Predict Y from X
Y_hat <- predict(fit_pls, X)
ok <- !is.na(Y_hat[, 1])
cat("Y prediction R-squared:",
    round(cor(Y[ok, 1], Y_hat[ok, 1])^2, 3), "\n")
```

## Projecting new data

All fitted objects support `project()` to transform new observations into the
latent space:

```{r project-new}
X_new <- matrix(rnorm(50 * m), 50, m)

# DiPCA / DiCCA
scores_new <- project(fit_pca, X_new)
dim(scores_new)
```

## Next steps

- `vignette("extracting-dynamic-components")` -- a deeper walkthrough with
  simulated ground-truth data, subspace recovery metrics, and method
  comparisons.
- `?dipca`, `?dicca`, `?dipls` -- full parameter documentation.
