---
title: "Extracting Dynamic Components from Multivariate Time Series"
author: "dipca package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting Dynamic Components from Multivariate Time Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Dynamic Inner PCA (DiPCA) and Dynamic Inner Canonical Correlation Analysis (DiCCA) are methods for extracting **autocorrelated latent components** from multivariate time series. Unlike standard PCA, which extracts static variance, these methods identify components that are both:

1. **Temporally structured** - Components have autocorrelation (they predict their own future)
2. **Orthogonal** - Components are mutually uncorrelated (like PCA)

This vignette demonstrates these methods on simulated data where we **know the ground truth**, making it crystal clear what the methods are recovering.

```{r libraries}
library(dipca)
library(multivarious)
library(ggplot2)
```

## The Ground Truth: Three Dynamic Latent Components

We'll simulate data with three hidden dynamic components, each with different temporal dynamics:

- **Component 1**: Strongly autocorrelated (ρ = 0.85) - slow oscillations
- **Component 2**: Moderately autocorrelated (ρ = 0.60) - medium dynamics
- **Component 3**: Weakly autocorrelated (ρ = 0.30) - fast changes

These latent components are then mixed through a loading matrix **P** to create 8 observed variables.

```{r simulate_data}
set.seed(123)

# Simulation parameters
N <- 500        # Number of time points
m <- 8          # Number of observed variables
h <- 3          # Number of latent components

# VAR(1) dynamics: T_t = A * T_{t-1} + noise
A <- diag(c(0.85, 0.60, 0.30))  # Autocorrelation strengths

# Generate latent components
T_true <- matrix(0, N, h)
for (t in 2:N) {
  T_true[t, ] <- A %*% T_true[t-1, ] + rnorm(h, sd = 0.3)
}

# Create orthogonal loading matrix (how latents mix to observables)
set.seed(456)
P_true <- qr.Q(qr(matrix(rnorm(m * h), m, h)))

# Observed data = latent components × loadings + noise
X <- T_true %*% t(P_true) + matrix(rnorm(N * m, sd = 0.2), N, m)

# Add column names for clarity
colnames(X) <- paste0("Var", 1:m)
colnames(T_true) <- paste0("True_", 1:h)
```

### Visualize the Ground Truth

```{r plot_ground_truth, fig.height=8}
# Plot true latent components
df_true <- data.frame(
  Time = rep(1:N, h),
  Component = rep(paste0("Component ", 1:h, " (ρ=", diag(A), ")"), each = N),
  Value = c(T_true[, 1], T_true[, 2], T_true[, 3])
)

p1 <- ggplot(df_true, aes(x = Time, y = Value)) +
  geom_line(color = "steelblue", size = 0.5) +
  facet_wrap(~ Component, ncol = 1, scales = "free_y") +
  labs(title = "Ground Truth: Three Dynamic Latent Components",
       subtitle = "Each component has different autocorrelation strength",
       y = "Latent Value") +
  theme_minimal()

# Plot observed variables (first 4 for clarity)
df_obs <- data.frame(
  Time = rep(1:N, 4),
  Variable = rep(paste0("Var", 1:4), each = N),
  Value = c(X[, 1], X[, 2], X[, 3], X[, 4])
)

p2 <- ggplot(df_obs, aes(x = Time, y = Value)) +
  geom_line(color = "darkgray", size = 0.3, alpha = 0.7) +
  facet_wrap(~ Variable, ncol = 1, scales = "free_y") +
  labs(title = "Observed Variables (First 4 of 8)",
       subtitle = "Mixtures of latent components plus noise",
       y = "Observed Value") +
  theme_minimal()

p1
p2
```

## Method 1: DiPCA (Dynamic Inner PCA)

DiPCA extracts components by maximizing their predictability from their own past while maintaining orthogonality.

```{r fit_dipca}
# Fit DiPCA with lag order s=1 (VAR(1) model), extract 3 components
fit_dipca <- dipca(X, s = 1, l = 3, n_init = 3, max_iter = 500,
                   preproc = center())

# Extract estimated components
T_dipca <- scores(fit_dipca)
colnames(T_dipca) <- paste0("DiPCA_", 1:3)
```

### How Well Did DiPCA Recover the Components?

```{r dipca_recovery, fig.height=8}
# Compare true vs estimated components
# Note: Components may be recovered in different order and with sign flips
# We'll compute subspace similarity

# Compute correlation between true and estimated subspaces
U_true <- qr.Q(qr(T_true[50:N, ]))    # Burn-in period
U_est  <- qr.Q(qr(T_dipca[50:N, ]))

# Subspace similarity (singular values of cross-product)
similarity <- svd(t(U_true) %*% U_est)$d

cat("Subspace recovery (singular values):\n")
print(round(similarity, 3))
cat("\nPerfect recovery would give [1, 1, 1]\n")
cat("We got:", paste(round(similarity, 3), collapse = ", "), "\n")

# Plot DiPCA components
df_dipca <- data.frame(
  Time = rep(1:N, 3),
  Component = rep(paste0("DiPCA Comp ", 1:3), each = N),
  Value = c(T_dipca[, 1], T_dipca[, 2], T_dipca[, 3])
)

ggplot(df_dipca, aes(x = Time, y = Value)) +
  geom_line(color = "firebrick", size = 0.5) +
  facet_wrap(~ Component, ncol = 1, scales = "free_y") +
  labs(title = "DiPCA Extracted Components",
       subtitle = paste0("Subspace recovery: ",
                        paste(round(similarity, 2), collapse = ", ")),
       y = "Component Value") +
  theme_minimal()
```

### DiPCA Properties: Orthogonality

One key property of DiPCA is that extracted components are **orthogonal** (uncorrelated).

```{r dipca_orthogonality}
# Compute correlation matrix of DiPCA components
cor_matrix <- cor(T_dipca)

# Visualize correlation matrix
df_cor <- as.data.frame(as.table(cor_matrix))
colnames(df_cor) <- c("Comp1", "Comp2", "Correlation")

ggplot(df_cor, aes(x = Comp1, y = Comp2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = round(Correlation, 2)), color = "white", size = 5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "DiPCA Component Correlations",
       subtitle = "Off-diagonal values should be near zero (orthogonal)") +
  theme_minimal() +
  coord_fixed()
```

### DiPCA Properties: Predictability

DiPCA components should be highly predictable from their own past. Let's check the one-step-ahead predictions.

```{r dipca_predictability, fig.height=5}
# Use predict() to get one-step-ahead forecasts
pred_dipca <- predict(fit_dipca, X)

# Compute R² for each component (correlation between t and t_hat)
r2_dipca <- sapply(1:3, function(j) {
  valid_idx <- !is.na(pred_dipca$scores_hat[, j])
  cor(pred_dipca$scores[valid_idx, j],
      pred_dipca$scores_hat[valid_idx, j])^2
})

cat("DiPCA component predictability (R²):\n")
cat("  Component 1:", round(r2_dipca[1], 3), "\n")
cat("  Component 2:", round(r2_dipca[2], 3), "\n")
cat("  Component 3:", round(r2_dipca[3], 3), "\n")

# Plot observed vs predicted for first component
df_pred <- data.frame(
  Time = 2:N,
  Observed = pred_dipca$scores[2:N, 1],
  Predicted = pred_dipca$scores_hat[2:N, 1]
)

ggplot(df_pred[1:100, ], aes(x = Time)) +
  geom_line(aes(y = Observed, color = "Observed"), size = 0.7) +
  geom_line(aes(y = Predicted, color = "Predicted"), size = 0.7, alpha = 0.7) +
  scale_color_manual(values = c("Observed" = "black", "Predicted" = "red")) +
  labs(title = "DiPCA Component 1: One-Step-Ahead Prediction",
       subtitle = paste0("R² = ", round(r2_dipca[1], 3)),
       y = "Component Value",
       color = "") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Method 2: DiCCA (Dynamic Inner Canonical Correlation)

DiCCA extracts components by maximizing the **canonical correlation** between each component and its own past. This typically yields components in descending order of predictability.

```{r fit_dicca}
# Fit DiCCA with lag order s=1, extract 3 components
fit_dicca <- dicca(X, s = 1, l = 3, inner = "classic", n_init = 3,
                   max_iter = 500, preproc = center())

# Extract estimated components
T_dicca <- scores(fit_dicca)
colnames(T_dicca) <- paste0("DiCCA_", 1:3)

# DiCCA stores R² values directly
cat("DiCCA component predictability (R²):\n")
cat("  Component 1:", round(fit_dicca$R2[1], 3), "\n")
cat("  Component 2:", round(fit_dicca$R2[2], 3), "\n")
cat("  Component 3:", round(fit_dicca$R2[3], 3), "\n")
cat("\nNote: DiCCA extracts components in descending R² order\n")
```

### DiCCA Property: Descending Predictability

DiCCA's canonical correlation objective ensures components are extracted in **descending order of predictability**.

```{r dicca_r2_plot}
df_r2 <- data.frame(
  Component = paste0("Comp ", 1:3),
  R2 = fit_dicca$R2,
  Method = "DiCCA"
)

ggplot(df_r2, aes(x = Component, y = R2)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = round(R2, 3)), vjust = -0.5, size = 4) +
  ylim(0, 1) +
  labs(title = "DiCCA: Components Ordered by Predictability",
       subtitle = "R² values should be monotonically decreasing",
       y = "R² (One-Step Prediction)") +
  theme_minimal()
```

### Compare DiPCA and DiCCA

```{r compare_methods, fig.height=6}
# Plot first components from both methods
df_compare <- data.frame(
  Time = rep(1:200, 2),
  Value = c(T_dipca[1:200, 1], T_dicca[1:200, 1]),
  Method = rep(c("DiPCA", "DiCCA"), each = 200)
)

ggplot(df_compare, aes(x = Time, y = Value, color = Method)) +
  geom_line(size = 0.6, alpha = 0.8) +
  scale_color_manual(values = c("DiPCA" = "firebrick", "DiCCA" = "steelblue")) +
  labs(title = "DiPCA vs DiCCA: First Component",
       subtitle = "Both methods extract similar dynamic structure",
       y = "Component Value") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Reconstruction: Going Back to Original Space

Both methods allow us to **reconstruct** the original data from the extracted components.

```{r reconstruction, fig.height=5}
# Reconstruct data using all 3 components
X_recon_dipca <- reconstruct(fit_dipca)
X_recon_dicca <- reconstruct(fit_dicca)

# Compute reconstruction error (RMSE)
rmse_dipca <- sqrt(mean((X - X_recon_dipca)^2))
rmse_dicca <- sqrt(mean((X - X_recon_dicca)^2))

cat("Reconstruction RMSE:\n")
cat("  DiPCA:", round(rmse_dipca, 4), "\n")
cat("  DiCCA:", round(rmse_dicca, 4), "\n")

# Plot original vs reconstructed for one variable
df_recon <- data.frame(
  Time = rep(1:200, 3),
  Value = c(X[1:200, 1], X_recon_dipca[1:200, 1], X_recon_dicca[1:200, 1]),
  Type = rep(c("Original", "DiPCA Recon", "DiCCA Recon"), each = 200)
)

ggplot(df_recon, aes(x = Time, y = Value, color = Type)) +
  geom_line(size = 0.5, alpha = 0.7) +
  scale_color_manual(values = c("Original" = "black",
                                 "DiPCA Recon" = "firebrick",
                                 "DiCCA Recon" = "steelblue")) +
  labs(title = "Original vs Reconstructed Data (Variable 1)",
       subtitle = paste0("DiPCA RMSE = ", round(rmse_dipca, 3),
                        ", DiCCA RMSE = ", round(rmse_dicca, 3)),
       y = "Value",
       color = "") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Dynamic Whitening: Removing Autocorrelation

A key application of DiPCA/DiCCA is **dynamic whitening** - removing temporal structure from the data. The residuals after reconstruction should have much less autocorrelation than the original data.

```{r whitening, fig.height=5}
# Compute residuals (prediction errors)
# Compute residuals (prediction errors) using class-specific method
resid_dipca <- dipca:::residuals.dipca(fit_dipca, X)

# Compute autocorrelation for original and residual data
acf_original <- mean(sapply(1:4, function(j) {
  acf(X[, j], lag.max = 1, plot = FALSE)$acf[2]
}))

acf_residual <- mean(sapply(1:4, function(j) {
  valid_idx <- is.finite(resid_dipca$e_hat[, j])
  if (sum(valid_idx) > 10) {
    acf(resid_dipca$e_hat[valid_idx, j], lag.max = 1, plot = FALSE)$acf[2]
  } else {
    NA
  }
}), na.rm = TRUE)

cat("Lag-1 Autocorrelation (average across variables):\n")
cat("  Original data:", round(acf_original, 3), "\n")
cat("  Residuals:    ", round(acf_residual, 3), "\n")
cat("  Reduction:    ", round(100 * (1 - abs(acf_residual)/abs(acf_original)), 1), "%\n")

# Visualize ACF reduction
df_acf <- data.frame(
  Type = c("Original", "Residuals"),
  ACF = c(acf_original, acf_residual)
)

ggplot(df_acf, aes(x = Type, y = ACF)) +
  geom_bar(stat = "identity", fill = c("gray50", "steelblue"), alpha = 0.7) +
  geom_text(aes(label = round(ACF, 3)), vjust = -0.5, size = 5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Dynamic Whitening Effect",
       subtitle = "DiPCA removes autocorrelation from residuals",
       y = "Lag-1 Autocorrelation",
       x = "") +
  theme_minimal()
```

## Summary

This vignette demonstrated:

1. **Simulation**: Created data with known ground truth - 3 dynamic latent components mixed into 8 observed variables

2. **DiPCA**:
   - ✓ Recovered the latent subspace (high similarity scores)
   - ✓ Extracted orthogonal components
   - ✓ Components are highly predictable from their past

3. **DiCCA**:
   - ✓ Extracted components in descending predictability order
   - ✓ Provided direct R² measures for each component

4. **Applications**:
   - Reconstruction: Go back to original variable space
   - Dynamic whitening: Remove autocorrelation from residuals
   - Forecasting: Predict future values using temporal structure

### When to Use Which Method?

- **DiPCA**: When you want orthogonal components that balance variance and predictability
- **DiCCA**: When you want components strictly ordered by predictability (canonical correlation)
- **DiCCA with ARIMA inner**: When your data has unit roots or complex ARMA dynamics

Both methods leverage the **multivarious** framework for consistent preprocessing and projection operations.

## References

Dong, Y., & Qin, S. J. (2018). Dynamic-inner canonical correlation and causality analysis for high dimensional time series data. *IFAC-PapersOnLine*, 51(18), 476-481.

## Session Info

```{r session_info}
sessionInfo()
```
